\LaTeXe\ guide for the Journal of Functional Programming

### 2019-05-02 papers/jfp/2-background.tex
Changes from SPJ
Minor edits in S2 and S3

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, to decide which files need to be rebuilt,
it uses the dependency graph that is dynamically constructed during
the previous build . This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting computations} -- see
Acar~\etal~\shortcite{acar2007selfadjusting} and~\S\ref{sec-related}. Second,
instead of aborting and deferring the execution of tasks whose newly discovered
dependencies have not yet been built (as \Excel does), \Shake \emph{suspends}
their execution until the dependencies are brought up to date. We refer to this
task scheduling algorithm as \emph{suspending}, see a concrete implementation
in~\S\ref{sec-implementation-shake}.

@@

\Shake also supports the \emph{early cutoff optimisation}. When it
executes a task and the result is unchanged from the previous build, it is
unnecessary to execute the dependent tasks, and hence \Shake can stop a build
earlier, as illustrated in Fig.~\ref{fig-cutoff}. Not all build systems support
early cutoff: \Shake and \Bazel (introduced below) do, but \Make and \Excel do
not.

@@

To support cloud builds, \Bazel maintains (i) a \emph{content-addressable cache}
that maps the hash of a file's content to the actual content of that file;
(ii) a memo table that records all executed build commands
with their input file hashes and output file value.
The memo table allows the build engine to bypass the execution
of a task, by predicting the hash of the result from the hashes of its
dependencies; then the content-addressable cache allows the engine
to download the result (if needed) based on the result hash. A concrete
implementation is provided in~\S\ref{sec-implementation-cloud}.

>>>>

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, to decide which files need to be rebuilt,
it uses the dependency graph that is dynamically constructed during
the previous build. This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting computations} -- see
Acar~\etal~\shortcite{acar2007selfadjusting} and~\S\ref{sec-related}. Second,
instead of aborting and deferring the execution of tasks whose newly discovered
dependencies have not yet been built (as \Excel does), \Shake \emph{suspends}
their execution until the dependencies are brought up to date. We refer to this
task scheduling algorithm as \emph{suspending}, see a concrete implementation
in~\S\ref{sec-implementation-shake}.

@@

\Shake also supports the \emph{early cutoff optimisation}. When it
executes a task and the result is unchanged from the previous build, it is
unnecessary to execute the dependent tasks, and hence \Shake can stop a build
earlier, as illustrated in Fig.~\ref{fig-cutoff}. Not all build systems support
early cutoff: \Shake and \Bazel (introduced below) do, but \Make and \Excel do
not (see \S\ref{sec-verifying-traces} for an explanation of why).

@@

To support cloud builds, \Bazel maintains (i)~a \emph{content-addressable cache}
that maps the hash of a file's content to the actual content of that file;
(ii)~a memo table that records all executed build commands
with their input and output file hashes.
The memo table allows the build engine to bypass the execution
of a task, by predicting the hash of the result from the hashes of its
dependencies; then the content-addressable cache allows the engine
to download the result (if needed) based on the result hash. A concrete
implementation is provided in~\S\ref{sec-implementation-cloud}.

### 2019-05-02 papers/jfp/3-abstractions.tex
Minor edits in S2 and S3

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Functor     f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Monad (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
runState  :: State s a -> s -> (a, s)
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty   -- mempty is identity for monoid m
    Const x <*> Const y = Const (x <> y) -- <> is the binary operation for m
\end{minted}
% \vspace{-3mm}
\caption{Standard library definitions.}\label{fig-stdlib}
% \vspace{-4mm}
\end{figure}

@@

Earlier we remarked that a \hs{Task}~\hs{Applicative} could only have static
dependencies. Usually we would extract such static dependencies by (in the case
of \Excel) looking at the syntax tree of the formula.  But a task description
has no such syntax tree: as you can see in the definition of \hs{Task} in Fig.~\ref{fig-types},
a task is just a function, so all we can do is call it.
\hs{Task}~\hs{Applicative} to find its dependencies \emph{without doing any of
the actual work}. Here is the code:

@@

So much for applicative tasks. What about monadic tasks with dynamic
dependencies? As we have seen in~\S\ref{sec-background-shake}, dynamic
dependencies need to be tracked too. This cannot be done statically; notice that
we cannot apply the function \hs{dependencies} to a \hs{Task}~\hs{Monad} because
the \hs{Const} functor has no \hs{Monad} instance. We need to run a monadic task
on a store with concrete values, which will determine the discovered
dependencies. Accordingly, we introduce the function \hs{track} -- a combination
of \hs{compute} and \hs{dependencies} that computes both the resulting value and
the list of its dependencies (key/value pairs) in an arbitrary monadic
context~\hs{m}:

@@

\noindent
This function uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad} for recording additional
information~--~a list of key/value pairs \hs{[(@@k,}~\hs{v)]}~--~when executing
a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch} with a
\hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key/value pair. The \hs{task} returns a value of type
\hs{WriterT}~\hs{[(@@k,}~\hs{v)]}~\hs{m}~\hs{v}, which we unwrap with
\hs{runWriterT}. We will use \hs{track} when implementing monadic build systems
with dynamic dependencies, see~\S\ref{sec-implementation-shake}. Here we show an
example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.

>>>>

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Functor     f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Monad (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
runState  :: State s a -> s -> (a, s)
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty   -- mempty is identity for monoid m
    Const x <*> Const y = Const (x <> y) -- <> is the binary operation for m
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Control.Monad.Trans.Writer
newtype WriterT w m a = WriterT { runWriterT :: m (a, w) }
tell :: Monad m => w   -> WriterT w m () -- write a value to the log
lift :: Monad m => m a -> WriterT w m a  -- lift an action into WriterT
\end{minted}

% \vspace{-3mm}
\caption{Standard library definitions.}\label{fig-stdlib}
% \vspace{-4mm}
\end{figure}

@@

Earlier we remarked that a \hs{Task}~\hs{Applicative} could only have static
dependencies. Usually we would extract such static dependencies by (in the case
of \Excel) looking at the syntax tree of the formula.  But a task description
has no such syntax tree: as you can see in the definition of \hs{Task} in Fig.~\ref{fig-types},
a task is just a function, so all we can do is call it.
Yet, remarkably, we can use the polymorphism of a
\hs{Task}~\hs{Applicative} to find its dependencies \emph{without doing any of
the actual work}. Here is the code:

@@

So much for applicative tasks. What about monadic tasks with dynamic
dependencies? As we have seen in~\S\ref{sec-background-shake}, dynamic
dependencies need to be tracked too. This cannot be done statically; notice that
we cannot apply the function \hs{dependencies} to a \hs{Task}~\hs{Monad} because
the \hs{Const} functor has no \hs{Monad} instance. We need to run a monadic task
on a store with concrete values, which will determine the discovered
dependencies. Accordingly, we introduce the function \hs{track} -- a combination
of \hs{compute} and \hs{dependencies} that computes both the resulting value and
the list of its dependencies (key/value pairs) in an arbitrary monadic
context~\hs{m}. We need this function to be polymorphic over~\hs{m}, because
each build system will execute tasks in its own monad, as we shall see
in~\S\ref{sec-implementations}.

@@

\noindent
This function uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad} for recording additional
information~--~a list of key/value pairs \hs{[(@@k,}~\hs{v)]}~--~when executing
a task in an arbitrary monad~\hs{m} (see the types in Fig.~\ref{fig-stdlib}).
We substitute the given \hs{fetch} with a
\hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key/value pair. The \hs{task} returns a value of type
\hs{WriterT}~\hs{[(@@k,}~\hs{v)]}~\hs{m}~\hs{v}, which we unwrap with
\hs{runWriterT}. We will use \hs{track} when implementing monadic build systems
with dynamic dependencies, see~\S\ref{sec-implementation-shake}. Here we show an
example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}, by defining
a corresponding \hs{fetchIO} of type \hs{String}~\hs{->}~\hs{IO}~\hs{Integer},
which allows us to demonstrate the dynamic nature of monadic dependencies in
GHCi.
